Reagents: expressing and composing fine-grained concurrency

### 2011-11-02 paper/paper.tex
related work, and new locking benchmarks

\authorinfo{}{}{}
% yikes!
%\title{Reagents: expressing and composing patterns of fine-grained concurrency}
\title{Reagents: expressing and composing fine-grained concurrency}
\maketitle

@@

Such libraries are an enormous undertaking---and one that must be repeated for
new platforms.  They tend to be conservative, implementing only those data
structures and primitives likely to fulfill common needs.  In addition, it is
generally not possible to safely combine the facilities of the library.  For
example, JUC provides queues, sets and maps, but not stacks or bags.  Its
queues come in both blocking and nonblocking forms, while its sets and maps
are nonblocking only.  Although the queues provide atomic (thread-safe)
dequeuing and sets provide atomic insertion, it is not possible to combine
these into a single atomic operation that moves an element from a queue into a
set.

@@

Broadly, we are interested in data structures and algorithms for
communication, synchronization, or both.  This section gives a brief
survey of the most important techniques for communication and
synchronization in a fine-grained setting---exactly the techniques
that reagents abstract and generalize.  Readers already familiar with
Treiber stacks~\cite{?}, elimination-backoff stacks~\cite{?}, dual
stacks~\cite{?}, and MCS locks~\cite{?} can safely skip to
\secref{reagents}.

@@

\elide{
class TreeBarrier {
  public readonly Synchronous.Channel[] Arrive;
  private readonly Join j; // create j, init chans ...
  public TreeBarrier(int n) {Wire(0, n-1, () => {});}
  private void Wire(int low, int high, Action Done) {
    if (low == high) j.When(Arrive[low]).Do(Done);
    else if (low + 1 == high)
      j.When(Arrive[low]).And(Arrive[high]).Do(Done);
    else { // low + 1 < high
      Synchronous.Channel Left, Right; // init chans       
      j.When(Left).And(Right).Do(Done);
      int mid = (low + high) / 2;
      Wire(low, mid, () => Left());
      Wire(mid + 1, high, () => Right()); 
    }
  }
}
}
\subsection{Dynamic reagents and \lstinline{cas}}

@@

Concurrent ML~\cite{Reppy1991} was designed to resolve an apparent tension
between abstraction and choice: if protocols are represented abstractly as
functions, it is impossible to express the choice of two abstract protocols.
The solution is \emph{higher-order concurrency}, in which synchronous
message-passing protocols are represented abstractly as \emph{events}.  CML's
events are built up from combinators, including a choice combinator,
communication combinators, and combinators for arbitrary computations not
involving communication.  Events are an abstract data type: externally to the
CML library, they appear as abstract, composable values.  Internally to the
library, however, events can be decomposed into their constituent combinators,
which allows CML to execute them.

Reagents are clearly influenced by the design of CML's events, and include
variants of CML's core event combinators (communication and choice).  Where
CML is aimed squarely at capturing synchronous communication protocols,
reagents are designed for writing and tailoring fine-grained concurrent data
structures and synchronization primitives.  This difference in motivation led
us to include a number of additional combinators, including those dealing
directly with shared state.

Originally, CML was focused on managing concurrency rather than profiting from
parallelism, and this focus was reflected in its implementation.  More
recently, a parallel implementation of CML was proposed~\cite{Reppy2009}.  The
key challenge is resolving uses of choice both consistently and scalably.  It
is addressed by \emph{sharing}: an event making a choice is enrolled as
offering a communication corresponding to each possible choice, and when a
communication is accepted, the (single, shared) event is atomically marked as
consumed.  We follow a similar strategy in dealing with message passing, but
where Parallel CML uses lock-based queues to store messages, we show how to
use lock-free bags for increased parallelism~(\secref{implementation}).  We
also show how to incorporate choice resolution with simultaneous shared-state
updates and our conjunction combinators.

@@

\subsection{Software transactional memory}

Software transactional memory was originally intended ``to provide a general
highly concurrent method for translating sequential object implementations
into non-blocking ones''~\cite{Shavit1997}.  This ambitious goal has led to a
remarkable research literature, which has been summarized in textbook
form~\cite{Larus2006}.  Much of the research is devoted to achieving
scalability on multiprocessors or multicores, sometimes by relaxing
consistency guarantees~\cite{?} or only providing obstruction-freedom rather
than lock-freemdom~\cite{?}.  %Work in the area is

@@

Reagents carve out a middle ground between completely hand-written algorithms
and the completely automatic \lstinline{atomic} blocks of STM.  When used in
isolation, reagents are \emph{guaranteed} to perform only the CASes that the
hand-written algorithm would, so they introduce no overhead on shared-memory
operations\elide{; we know of no STM implementation that can provide similar
  guarantees or a similar cost model}.  Yet, unlike hand-written algorithms,
reagents can be tailored and composed using choice, adding blocking behavior,
or combining sequentially into larger atomic blocks. %TODO

Haskell's STM~\cite{Harris2005a}

left-biased choice, but differently from \lstinline{orElse} in
Haskell's STM.

@@

Our mixture of message-passing and shared-state combinators most closely
resembles the work on Communicating Transactions.   %TODO

@@

\subsection{Composing lock-free data structures}

Most of the literature on fine-grained concurrent data structures studies only
``within-object'' atomicity, for example developing algorithms for inserting
or removing elements into a collection atomically.  However, there has been
some nice work studying the problem of transferring data atomically between
such fine-grained data structures~\cite{Cederman2010}.  The basic approach
relies on a \emph{k}CAS operation in much the same way that reagent sequencing
does.  The transfer methods must be defined manually, in advance, and with
access to the internals of the relevant data structures, whereas reagents
allow arbitrary new compositions, without manual definition, and without
access to the code or internals of the involved data structures.

Atomic transfers between concurrent data structures

elastic transactions~\cite{Felber2009} and company,

composing lock-free data structure operations~\cite{Cederman2010}

>>>>

\authorinfo{}{}{}
%\authorinfo{Aaron Turon}{Northeastern University}{turon@ccs.neu.edu}

\title{Reagents: expressing and composing fine-grained concurrency}
\maketitle

@@

Such libraries are an enormous undertaking---and one that must be repeated for
new platforms.  They tend to be conservative, implementing only those data
structures and primitives likely to fulfill common needs, and it is generally
not possible to safely combine the facilities of the library.  For example,
JUC provides queues, sets and maps, but not stacks or bags.  Its queues come
in both blocking and nonblocking forms, while its sets and maps are
nonblocking only.  Although the queues provide atomic (thread-safe) dequeuing
and sets provide atomic insertion, it is not possible to combine these into a
single atomic operation that moves an element from a queue into a set.

@@

\cite{Shavit1997a} %TODO: cite pure elimination

Broadly, we are interested in data structures and algorithms for
communication, synchronization, or both.  This section gives a brief survey of
the most important techniques for communication and synchronization in a
fine-grained setting---exactly the techniques that reagents abstract and
generalize.  Readers already familiar with Treiber stacks~\cite{Treiber1986},
elimination-backoff stacks~\cite{Hendler2004}, dual stacks~\cite{Scherer2004},
and MCS locks~\cite{Mellor-Crummey1991} can safely skip to \secref{reagents}.

@@

\elide{
class TreeBarrier {
  public readonly Synchronous.Channel[] Arrive;
  private readonly Join j; // create j, init chans ...
  public TreeBarrier(int n) {Wire(0, n-1, () => {});}
  private void Wire(int low, int high, Action Done) {
    if (low == high) j.When(Arrive[low]).Do(Done);
    else if (low + 1 == high)
      j.When(Arrive[low]).And(Arrive[high]).Do(Done);
    else { // low + 1 < high
      Synchronous.Channel Left, Right; // init chans       
      j.When(Left).And(Right).Do(Done);
      int mid = (low + high) / 2;
      Wire(low, mid, () => Left());
      Wire(mid + 1, high, () => Right()); 
    }
  }
}
}
\subsection{Dynamic reagents and \lstinline{cas}}
\label{sec:dynamic}

@@

Concurrent ML~\cite{Reppy1991} was designed to resolve an apparent tension
between abstraction and choice: if protocols are represented abstractly as
functions, it is impossible to express the choice of two abstract protocols.
The solution is \emph{higher-order concurrency}, in which synchronous
message-passing protocols are represented abstractly as \emph{events}.  CML's
events are built up from combinators, including a choice combinator,
communication combinators, and combinators for arbitrary computations not
involving communication.  \elide{
Events are an abstract data type: externally to the
CML library, they appear as abstract, composable values.  Internally to the
library, however, events can be decomposed into their constituent combinators,
which allows CML to execute them.
}
Reagents are clearly influenced by the design of CML's events, and include
variants of CML's core event combinators\elide{(communication and choice)}.
But where CML is aimed squarely at capturing synchronous communication
protocols, reagents are designed for writing and tailoring fine-grained
concurrent data structures and synchronization primitives.  This difference in
motivation led us to include a number of additional combinators, including
those dealing directly with shared state.

Originally, CML was focused on managing concurrency rather than profiting from
parallelism, and this focus was reflected in its implementation.  More
recently, a parallel implementation of CML was proposed~\cite{Reppy2009}.  The
key challenge is resolving uses of choice both consistently and scalably.  It
is addressed by \emph{sharing}: an event making a choice is enrolled as
offering a communication corresponding to each possible choice, and when a
communication is accepted, the (single, shared) event is atomically marked as
consumed.  We follow a similar strategy in dealing with message passing, but
where Parallel CML uses lock-based queues to store messages, we show how to
use lock-free bags for increased parallelism~(\secref{implementation}).  We
also show how to incorporate choice resolution with shared-state updates and
our conjunction combinators.

@@

\subsection{Software transactional memory}
\label{sec:stm}

Software transactional memory was originally intended ``to provide a general
highly concurrent method for translating sequential object implementations
into non-blocking ones''~\cite{Shavit1997}.  This ambitious goal has led to a
remarkable research literature, which has been summarized in textbook
form~\cite{Larus2006}.  Much of the research is devoted to achieving
scalability on multiprocessors or multicores, sometimes by relaxing
consistency guarantees~\cite{?} or only providing obstruction-freedom rather
than lock-freemdom~\cite{Herlihy2003}.  %Work in the area is

@@

Reagents carve out a middle ground between completely hand-written algorithms
and the completely automatic \lstinline{atomic} blocks of STM.  When used in
isolation, reagents are \emph{guaranteed} to perform only the CASes that the
hand-written algorithm would, so they introduce no overhead on shared-memory
operations; by recoding an algorithm use reagents, you lose nothing.\elide{;
  we know of no STM implementation that can provide similar guarantees or a
  similar cost model} Yet unlike hand-written algorithms, reagents can be
composed using choice, tailored with new blocking behavior, or combined
into larger atomic blocks. %TODO

Haskell's STM~\cite{Harris2005a} was inspirational in showing that
transactions can be represented via monads~\cite{PeytonJones1993}, explicitly
composed, and combined with blocking and choice operators.  Reagents also form
a monad, but we have chosen an interface closer to
\emph{arrows}~\cite{Hughes2000}, to encourage static reagent layout wherever
possible~(\secref{dynamic}).  Like \lstinline{orElse} in Haskell's STM, our
choice operator is left-biased.  But unlike \lstinline{orElse}, our choice
operator will attempt the right-hand side even when the left-hand side has
only failed transiently (rather than permanently).  While the distinction may
appear somewhat technical, it is crucial for supporting examples like the
elimination backoff stack~(\secref{choice}).

@@

Our mixture of message-passing and shared-state combinators most closely
resembles the work on Communicating Transactions.  Of course, the most
important difference is in the way we deal with shared state (and the
guarantees provided thereby), which we discussed in~\secref{stm}.  Aside from
that, we believe that \emph{synchronous} communication is a better choice for
expressing patterns like elimination~(\secref{choice}), since the pattern
relies on both participants being mutually-aware.  %TODO: say more here

@@

\subsection{Composing fine-grained concurrent data structures}

Most of the literature on fine-grained concurrent data structures is focused
on ``within-object'' atomicity, for example developing algorithms for
inserting or removing elements into a collection atomically.  However, there
has been some recent work studying the problem of transferring data atomically
between such fine-grained data structures~\cite{Cederman2010}.  The basic
approach relies on a \emph{k}CAS operation in much the same way that reagent
sequencing does.  The transfer methods must be defined manually, in advance,
and with access to the internals of the relevant data structures, whereas
reagents allow arbitrary new compositions, without manual definition, and
without access to the code or internals of the involved data structures.
Nevertheless, the implementation of reagent composition yields an algorithm
very similar to the manually-written transfer methods.

It is also possible to go in the other direction: start from STM, which
provides composition, and add an ``escape hatch'' for writing arbitrary
fine-grained concurrent algorithms within the scope of a transaction.  The
escape hatch can be provided through unlogged reads and/or writes to memory
locations being used by transactions, as in early release~\cite{Herlihy2003}
or elastic transactions~\cite{Felber2009}.  As we discussed
above~(\secref{stm}), we favor an approach where the focus is foremost on
writing fine-grained algorithms, with \emph{guarantees} about the performance
and shared-memory semantics of those algorithms.  Providing such guarantees
via an escape hatch mechanism may be difficult or impossible, depending on the
details of the STM implementation.  As we showed in~\secref{reagents}, it is
also very useful to have combinators for choice, message-passing, and
blocking, if one wishes to capture the full range of fine-grained concurrent
algorithms.
